\documentclass[12pt,letterpaper,noanswers]{exam}
\usepackage[margin=0.9in]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{multicol}
\pagestyle{head}
\header{AM 111}{}{Preliminary Project Proposal}
\runningheadrule
\headrule
\usepackage{graphicx} % more modern
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{hyperref}
\usepackage{tcolorbox}

\begin{document}
 \pdfpageheight 11in 
  \pdfpagewidth 8.5in

The goal of the final project is for you to apply numerical analysis or scientific computing techniques to an applied or theoretical problem, or to conduct an investigation into a method not discussed in class.

\begin{enumerate}
\itemsep0pt
    \item The (short) preliminary project proposals are due on Thursday at 5pm.
    \item Each individual group member is responsible for submitting a different one, so your group will propose as many projects as the group has members.
    \item Because these papers are popular, your group can propose either the flight reconstruction paper or the deep learning paper, but not both (as one of your three proposals).
    \item Each team will work on a different project (with only one team per paper or resource).
\end{enumerate}


\noindent For each preliminary project, answer the following questions:
\begin{itemize}
\itemsep0pt
\item What will be the aim or purpose of the project?
\item What paper(s) or resource(s) would you like to base your project work around?
\item How does the intended project incorporate mathematical topics related to this course?
\item What method or methods do you intend to implement as part of the project?
\item What do you think you will you need to learn to be able to complete your project?
\end{itemize}

\noindent Project submissions:
\begin{itemize}
\item Due Thursday Oct 27 at 5pm: Multiple preliminary proposals (individual or group submission)
    \item Due Friday Nov 4: Project proposal for assigned project topic (group submission).  Timeline of project work and deliverables (individual submission).
    \item Due Friday Nov 11: Weekly project update and work log (individual submission)
    \item Due Friday Nov 18: Weekly project update and work log (individual submission).  Draft presentation slides.  Minimum of 1 slide per team member, maximum of 2 slides per team member.  (individual submission)
    \item Due Friday Nov 25: Weekly project update and work log (individual submission).  
    \item Tuesday Nov 29 or Thurs Dec 1: Presentation explaining your project topic to the class (group submission).  1 person team: 2 minutes.  2 person team: 3.5 minutes.  3 person team: 4.5 minutes.
    \item Due Friday Dec 2: Weekly project update and work log (individual submission).  Draft final project summary (group submission).
    \item Due Wednesday Dec 7: Peer review on final project summary (individual submission)
    \item Due Friday Dec 9: Weekly project update and work log (individual submission).
    \item Due Thursday Dec 15: Final project summary (group submission).  Individual project report (individual submission).
\end{itemize}

\eject
\subsection*{Project Groups}
\begin{multicols}{2}
\begin{enumerate}
\itemsep0pt
    \item Aidan, Alex B, Basil
    \item Alex H, Eric, Ivonne
    \item Benjamin, Eletria, Mack
    \item Brian, Emma, Ray
    \item Caitlin, Dani, Julia K
    \item Cameron, Kevin C, Robert
    \item Daniyal, Eli, Zachary
    \item Esmé, Jessica, Kevin G
    \item Jack, Mina, Tom
    \item Johan
    \item Julia M, Nicolas, Sophie
    \item Marissa, Shang
    \item Nina, Nini, RJ
\end{enumerate}
\end{multicols}

\end{document}