\documentclass[12pt,letterpaper,noanswers]{exam}
%\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[margin=0.9in]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{multicol}
\pagestyle{head}
\definecolor{c02}{HTML}{FFBBBB}
\definecolor{c03}{HTML}{FFDDDD}
\header{AM 111 Third Reflection}{}{} %Updated on \today.}{}
\runningheadrule
\headrule
\usepackage{diagbox}
\usepackage{graphicx} % more modern
%\usepackage{subfigure} 
\usepackage{amsmath} 
\usepackage{amssymb} 
%\usepackage{gensymb} 
%\usepackage{natbib}
\usepackage{hyperref}
%\usepackage{enumitem}
%\setlength{\parindent}{0pt}
%\usepackage{setspace}
%\pagestyle{empty}  
%\newcommand{\Sc}[0]{
%{\color{BlueViolet}\S}
%}
\usepackage{tcolorbox}
\usepackage[framed,numbered,autolinebreaks,useliterate]{mcode}

\begin{document}
 \pdfpageheight 11in 
  \pdfpagewidth 8.5in
\noindent  Name:
\vspace{0.2cm}

\noindent Revise and update your responses to your previous reflection

\begin{enumerate}
    \item At this time, what are your goals for yourself in this course?
    \vspace{1.7in}
    
\item Articulate your current plan for how you will “do” the rest of this course. In your plan, please include how much time you will allocate to this course each week, what you will do when you encounter challenges, and so forth. Be as specific and granular as you can.  Include information about the final project.

\vspace{3in}
\eject 
\item Keeping in mind the goals above, what specific criteria / metrics / experiences / reflections will you use to evaluate yourself? In other words, how will you know if you are meeting your goals? \emph{I am not asking you to evaluate yourself yet. I am asking you to come up with a specific plan for how you will evaluate yourself.}
\vspace{3in}

\item Now, using your response to the previous question, please evaluate your work in this course so far.
\vspace{2.5in}
\item You'll be revising your answers to this self-evaluation form once more. In light of your own evaluation of yourself, is there anything in your approach to this course that you will plan to change between now and the next time you fill out this form? If so, please explain.
\end{enumerate}


\end{document}
