\documentclass[12pt,letterpaper,noanswers]{exam}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[margin=0.9in]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{multicol}
\pagestyle{head}
\header{AM 111}{}{Final Reflection p.\thepage}
\runningheadrule
\headrule
\usepackage{graphicx} % more modern
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{hyperref}
\usepackage{tcolorbox}

\begin{document}
 \pdfpageheight 11in 
  \pdfpagewidth 8.5in

\noindent 




%\noindent Upload your reflection to Gradescope by May 8th at 11:59pm.

\begin{enumerate}
    \item Why did you take this course? What were your goals for yourself?
    \item How did you approach this course? How much time did you spend on it each week? What did you do when you encountered challenges? Please include these and any other relevant details.
    \item Keeping in mind the goals you articulated above, what specific criteria / metrics / experiences / reflections are you using to evaluate yourself? In other words, how will you know if you have met your goals? [Note: I am not asking you to evaluate yourself yet. I am asking you to describe your final plan for evaluating yourself.]
    \item Now, using your response to the previous question, please evaluate your work in this course overall. This evaluation should be narrative. I'll ask you for a letter grade at the end of this reflection.
    \item Read the Harvard College grading system information available here: \url{https://infoforfaculty.fas.harvard.edu/book/grading-system}. This may be a different set of criteria/standards than the ones you generated in question 3.  It is focused on comprehension of the course material, command of the skills needed to work with the course material, and level of engagement with the course requirements and activities.  Provide a narrative evaluation of your work in the course overall using these criteria.  I will ask you for a second letter grade at the end of this reflection.
    \item If you had this course to do over again, assuming the course itself stayed the same, would you change anything about how you approached it? If so, what?
    \item How, if at all, did this course affect your feelings about mathematics?
    \item Do you plan to do anything mathematical in the future? (Courses, reading, research, bringing math into other fields you study... anything.)
    \item You will certainly have a thorough chance to evaluate this course on the usual anonymous end-of-term course evaluation. That said, if there is anything you want to tell me here, please feel free!
    \item What letter grade should I assign you based on the criteria in question 3 and your narrative evaluation in question 4?
    \item What letter grade would be assigned based on the criteria and evaluation in question 5?  
    
    Note that the grade of 'A' requires your work in the course to be ``of extraordinary distinction'', which is very rare.  The grade of 'A-' would reflect excellent quality work and full mastery of the subject, which is also a very high standard.
    
    \emph{An inflated assessment of your level of mastery / work quality may lead to a small deduction in your assigned letter grade.}
    
    
    \item The grades in question 10 and 11 may be different.  If that is the case, what letter grade should I assign you?
\end{enumerate}










\end{document}